---
title: A Practical Guide to Retrieving SSURGO Soil Data in R
authors:
  - skazemi
categories:
  - R
  - Soil Data
  - SSURGO
draft: true
date: "2025-06-06T00:00:00Z"
image:
  caption: ""
  focal_point: ""
lastMod: "2025-06-06T00:00:00Z"
subtitle: How to query, process, and summarize soil data from SSURGO in R.
summary: "This detailed tutorial walks through how to retrieve and summarize SSURGO soil data using spatial coordinates in R. It includes custom function writing and automation over multiple locations."
tags: []
---

Working with agricultural data often requires accurate soil information. Instead of manually searching soil survey data, we can access the **SSURGO** (Soil Survey Geographic Database) through the `soilDB` package in R using spatial queries.

This guide shows how to:
- Query the SSURGO database for a specific point
- Retrieve soil attributes like texture, bulk density, Soil series classification, Soil depth,soil organic matter and Water holding capacity 
- Write a reusable function
- Apply it to a whole dataset
---



### Required Packages

We’ll need the following R packages installed and loaded:
```{r setup, include=FALSE}
library(FedData)
library(dplyr)
library(sf)
library(readr)
library(soilDB)
```
These libraries enable querying soil databases (soilDB), manipulating data (dplyr), and handling spatial formats (sf).


#Step 1: Get Soil Data for a Single Point
Start with a test location. We'll use latitude and longitude to generate a spatial point in Well-Known Text (WKT) format and query the SDA (Soil Data Access) API for its MUKEY — a key that uniquely identifies a soil map unit.
```{r}
# Coordinates for the example point
lon <- -97.0
lat <- 47.0
my_point <- sprintf("POINT(%f %f)", lon, lat)

# Get MUKEY (Map Unit Key)
sql <- sprintf("SELECT mukey FROM SDA_Get_Mukey_from_intersection_with_WktWgs84('%s')", my_point)
mukey_result <- SDA_query(sql)
mukey <- mukey_result$mukey[1]
print(mukey)
```
This code:
- Formats the point as WKT (Well-Known Text)
- Sends a SQL query to the Soil Data Access API
- Extracts the Map Unit Key (`mukey`) for the location
- What does mukey[1] mean?

When we query the SSURGO database using spatial coordinates, the result may include more than one map unit (MUKEY) — especially if the point falls on the border of two or more soil map units. The result is returned as a data frame To keep things simple, we just select the first MUKEY using: 
mukey <- mukey_result$mukey[1].



#Step 2: Retrieve Soil Properties by MUKEY
Once we have the `mukey`, we can extract soil attributes using another SQL query.
```{r}
sql_comp <- sprintf("
  SELECT c.mukey, c.compname, c.comppct_r, ch.hzdept_r, ch.hzdepb_r,
         ch.awc_r, ch.om_r, ch.dbthirdbar_r
  FROM component c
  INNER JOIN chorizon ch ON c.cokey = ch.cokey
  WHERE c.mukey = '%s'", mukey)
```
Once we have the mukey, we can request detailed information about the soil properties associated with that map unit. This involves two steps: first, we write a SQL query to retrieve data from multiple SSURGO tables.

component table (c):

Holds general information about each soil component (e.g. name, percent of the map unit, key ID).

compname: soil series name

comppct_r: the percentage this component contributes to the map unit

chorizon table (ch):

Contains detailed data about individual soil horizons (layers) within each component.

hzdept_r and hzdepb_r: top and bottom depths of each horizon

awc_r: available water capacity (how much water the soil can store)

om_r: organic matter content

dbthirdbar_r: bulk density (mass per volume, dry weight)

INNER JOIN ch ON c.cokey = ch.cokey:

Links the component table to the horizon table using cokey, the component key.

This allows us to pull horizon-level data for each component.

WHERE c.mukey = '%s':

Filters the results so we only get data for the specified map unit key (mukey).

The result is a long data frame where each row represents a single horizon from a component within the selected map unit.


```{r}
soil_properties <- SDA_query(sql_comp)

soil_summary <- soil_properties %>%
  group_by(mukey) %>%
  summarise(
    soil_series = first(compname),
    soil_depth_cm = max(hzdepb_r, na.rm = TRUE),
    awc = round(mean(awc_r, na.rm = TRUE), 2),
    organic_matter = round(mean(om_r, na.rm = TRUE), 2),
    bulk_density = round(mean(dbthirdbar_r, na.rm = TRUE), 2),
    .groups = "drop"
  )

print(soil_summary)
```
then we summarize the results in R to make them easier to interpret.

group_by(mukey):

Groups the data by map unit (though usually there's only one in this context).

soil_series = first(compname):

Uses the first component name as a representative "soil series."

we could refine this by choosing the most dominant component (e.g., with the highest comppct_r), but here we keep it simple.

soil_depth_cm = max(hzdepb_r):

Calculates the total depth by finding the bottom of the deepest horizon.

awc, organic_matter, bulk_density:

These are averaged across all horizons to create a single summary value per map unit.

round() is used to make the output cleaner (e.g., 1.47 instead of 1.472365).

.groups = "drop":

Ensures that the result is a flat data frame (not a grouped tibble).

This step transforms multiple horizons into one easy-to-use row of summary metrics.

#Step 3: Automate Soil Data Retrieval with a Custom Function 
We can now write a custom function that can take any lat/lon and return a row of soil data, rather than repeating the same SSURGO query logic for every location manually. This makes it easy to retrieve soil data for many locations in a clean and reusable way.
```{r}
get_soil_info_from_coords <- function(trial_id, lon, lat) {
  tryCatch({
    wkt <- sprintf("POINT(%f %f)", lon, lat)

    sql_mukey <- sprintf(
      "SELECT mukey FROM SDA_Get_Mukey_from_intersection_with_WktWgs84('%s')", wkt
    )
    mukey_result <- SDA_query(sql_mukey)
    if (nrow(mukey_result) == 0) return(NULL)

    mukey <- mukey_result$mukey[1]

    sql_comp <- sprintf("
      SELECT c.mukey, c.compname, ch.chkey, ch.hzdept_r, ch.hzdepb_r,
             ch.awc_r, ch.om_r, ch.dbthirdbar_r, tg.texture
      FROM component c
      INNER JOIN chorizon ch ON c.cokey = ch.cokey
      LEFT JOIN chtexturegrp tg ON ch.chkey = tg.chkey
      WHERE c.mukey = '%s'", mukey)

    soil_props <- SDA_query(sql_comp)
    if (nrow(soil_props) == 0) return(NULL)

    surface_texture <- if ("texture" %in% names(soil_props)) {
      soil_props %>%
        filter(hzdept_r == min(hzdept_r, na.rm = TRUE)) %>%
        pull(texture) %>% .[1]
    } else { NA_character_ }

    summarise(soil_props,
      trial_id = trial_id,
      soil_series = first(compname),
      surface_texture = surface_texture,
      soil_depth_cm = max(hzdepb_r, na.rm = TRUE),
      awc = round(mean(awc_r, na.rm = TRUE), 2),
      organic_matter = round(mean(om_r, na.rm = TRUE), 2),
      bulk_density = round(mean(dbthirdbar_r, na.rm = TRUE), 2)
    )
  }, error = function(e) {
    message(sprintf("Error processing trial %s: %s", trial_id, e$message))
    return(NULL)
  })
}
```

How this function works:
Inputs: It takes three inputs: a unique trial_id, and the geographic coordinates (lon, lat) for a sampling or trial location.

WKT Conversion: It converts the lat/lon to a spatial format ("POINT(lon lat)") understood by the SSURGO API.

MUKEY Query: It retrieves the map unit key (mukey) that corresponds to that location.

Soil Properties Query: It fetches detailed soil horizon information for the mukey from SSURGO’s component and horizon tables.

Surface Texture: If texture data exists, it extracts the surface horizon’s texture type (the shallowest layer).

Summarization: It calculates summary statistics across horizons, including:

Total soil depth (hzdepb_r)

Mean available water capacity (awc_r)

Organic matter content (om_r)

Bulk density (dbthirdbar_r)

Output: The function returns a single-row data frame with all relevant soil information for one location.

Error Handling: If anything goes wrong (e.g., no mukey found, or failed query), it gracefully returns NULL and prints a helpful message.

By encapsulating this entire workflow into a function, we make it easy to apply the same logic to many locations using a simple loop or lapply().



# Step 4: Apply to Real Metadata(Dataset of Locations)
Assume we have a dataset `my_data` with `trial_id`, `trial_latitude`, and `trial_longitude`.

```{r}
head(my_data)
```
  trial_id   trial_latitude   trial_longitude
  T001       47.12            -97.45
  T002       47.35            -97.20
  ...


We can now apply the get_soil_info_from_coords() function to each row:
```{r}
soil_results_list <- lapply(1:nrow(my_data), function(i) {
  row <- my_data[i, ]
  get_soil_info_from_coords(
    trial_id = row$trial_id,
    lon = row$trial_longitude,
    lat = row$trial_latitude
  )
})

# Combine all the rows of results
soil_data <- do.call(rbind, soil_results_list)
```

# Step 5: Merge with Trial Data
Now that we've gathered soil metrics for each location, merge it back with our original dataset:
```{r}
my_data_with_soil <- my_data %>%
  left_join(soil_data, by = "trial_id")
```
Now our trial data includes matched soil metrics like texture, AWC, OM, and bulk density.


# Summary

This tutorial covered how to:
- Query SSURGO using SDA and coordinates
- Extract soil variables from SQL
- Write a robust, reusable R function
- Apply it to real-world trial data

By wrapping the logic in a function, we save time and reduce errors in repeated data extraction tasks.

This method is especially useful for agronomic field trials, ecological sampling, or geospatial modeling projects.
---

## Simulated Soil Data for Visualization
 Just want to try the visualizations quickly, here’s a sample dataset with realistic values based on SSURGO outputs.
```{r}
soil_data <- tibble::tibble(
  trial_id = c("T001", "T002", "T003", "T004", "T005"),
  trial_latitude = c(47.12, 47.35, 47.55, 47.25, 47.45),
  trial_longitude = c(-97.45, -97.20, -97.30, -97.15, -97.40),
  soil_series = c("Barnes", "Fargo", "Bearden", "Barnes", "Fargo"),
  surface_texture = c("silty clay loam", "clay loam", "loam", "silt loam", "clay loam"),
  soil_depth_cm = c(150, 180, 120, 160, 170),
  awc = c(0.15, 0.17, 0.13, 0.16, 0.18),
  organic_matter = c(3.2, 4.0, 2.8, 3.5, 3.8),
  bulk_density = c(1.3, 1.2, 1.4, 1.3, 1.25)
)
```

# Step 6: Visualizing Simulated Soil Data


#### 1. Barplot of Soil Series by Trial count
```{r}
ggplot(soil_data, aes(x = soil_series)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Number of Trials per Soil Series",
       x = "Soil Series",
       y = "Trial Count") +
  theme_minimal()

```

#### 2. Point Plot of AWC
```{r}
ggplot(soil_data, aes(x = trial_id, y = awc)) +
  geom_point(size = 3, color = "steelblue") +
  labs(title = "AWC by Trial ID", x = "Trial ID", y = "AWC (cm/cm)") +
  theme_minimal()
```

#### 3. Boxplot of Soil Depth by Soil Series
```{r}
ggplot(soil_data, aes(x = soil_series, y = soil_depth_cm)) +
  geom_boxplot() +
  labs(title = "Soil Depth by Soil Series",
       x = "Soil Series",
       y = "Depth (cm)") +
  theme_minimal()
```
#### 4. Boxplot of Organic Matter Distribution
```{r}
ggplot(soil_data, aes(x = soil_series, y = organic_matter)) +
  geom_boxplot() +
  labs(title = "Organic Matter by Soil Series",
       x = "Soil Series",
       y = "Organic Matter (%)") +
  theme_minimal()
```
#### 5. Surface Texture Frequency
```{r}
soil_data %>%
  count(surface_texture) %>%
  ggplot(aes(x = reorder(surface_texture, -n), y = n)) +
  geom_col(fill = "coral") +
  labs(title = "Surface Texture Distribution",
       x = "Surface Texture",
       y = "Number of Trials") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


---

# Step 6: Mapping Soil Variables

Let’s map our soil data spatially to see where different conditions exist.

#### 1. Map: Soil Depth at Trial Locations
For this example, we’ll use the built-in maps package to draw a basic background map of a region. We can customize the region argument to fit our study area.
```{r}
library(maps)
base_map <- map_data("state", region = c("north dakota", "minnesota"))

ggplot() +
  geom_polygon(data = base_map,
               aes(x = long, y = lat, group = group),
               fill = "white", color = "gray") +
  geom_point(data = soil_data,
             aes(x = trial_longitude, y = trial_latitude, color = soil_series),
             size = 4, alpha = 0.8) +
  coord_fixed(1.3) +
  labs(title = "Trial Locations by Soil Series",
       x = "Longitude", y = "Latitude") +
  theme_minimal()
```

These visualizations help us explore how soil properties vary across geography and between trial sites.
